Stage-Splitting a Modal Language

### 2014-10-16 paper/example.tex
clarifications to section two, and a new word for splitting residuals

To be precise, our coloring strategy respects data dependence, but not control dependence---the 
first arguments to the recursive calls to \texttt{qSelect} are colored red,
despite appearing underneath a blue case statement.
Because of this coloring strategy and the definition of data dependence, 
information can flow from red code to blue code but not vice versa.

Observe the that helper function \texttt{partition} is colored entirely red.
This is where much of the work of the quickselect algorithm takes place, and it doesn't even depend on $k$!

This suggests an optimization: we can split \texttt{qSelect} into two functions, 
one for all of the red code and another for all of the blue code. 
We'll call them \texttt{qSelect1} and \texttt{qSelect2}.
Because of the dependence properties stated above, 
\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
whereas \texttt{qSelect1} will only depend on the input list $l$.
Of course, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
Equationally, we can represent this relationship as simply 
\[
\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
\]

With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from \texttt{qSelect}.
Because the original function is recursive,
the first step in this process is to assume this conclusion at the call sites.
This is performed in place:
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}
We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}
This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
We continue to pull up the precomputation and include more in it:

>>>>

To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
first arguments to the recursive calls to \texttt{qSelect} are colored red,
despite appearing underneath a blue case statement.
This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.

Observe the that helper function \texttt{partition} is colored entirely red,
which is valid because it is called with only red arguments.  
This is an illustration of the property stated above, that 
the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.

This suggests an optimization: we can split \texttt{qSelect} into two functions, 
one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
Because of the dependence properties stated above, 
\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
whereas \texttt{qSelect1} will only depend on the input list $l$.
Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
Equationally, we can represent this relationship as simply 
\[
\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
\]
for some list \texttt{l} and index \texttt{k}.

With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
Because the original function is recursive,
the first step in this process is to assume the conclusion at the call sites.
This is performed in place:
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}
We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}
This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
We continue to pull up the precomputation and include more in it:

### 2014-10-16 paper/splitting.tex
clarifications to section two, and a new word for splitting residuals

\ur{The reuse of residual is bothersome, we should use different terminology,
1-value and 2-value perhaps.}

Splitting arbitrary terms in \lang\ again results in two terms containing only
the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
called, respectively, the {\em precomputation} and {\em residual}.%
\footnote{We also used the term {\em residual} for the result of
partial evaluation.  Where necessary, we will distinguish between
these concepts by using the terms {\em splitting-residual} and {\em
evaluation-residual}.}
%
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a residual
\verb|fn (L1,L2) => L1 < L2-5|, with the property that the residual applied to
the precomputation must agree with the original term. This process of hoisting
all the stage \bbone\ subterms out of a term is called \emph{lifting}.

@@

The precomputation for this term is a \emph{combined result} which is
the pair of these components: \verb|(((),3*4),1+2)|. The splitting residual is
likewise the pair of the stage \bbtwo\ results of lifting and masking:
\verb|fn L => (L<5,())|.
%\footnote{In this example, the combined result is a 2-tuple, but more generally
%it only must be a term which reduces to a 2-tuple.}

@@

The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
$c$ and residual $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
precomputation $p$ and a residual $l.r$. In both cases, the term $e$ is open on
variables in the context $\Gamma$. We write the rules using pattern notation
when appropriate.
%For concision, we represent residuals as a term open on a single variable,
%rather than as functions.

@@

The $\splittwosym$ rules are conceptually simple and almost all very similar.  
As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
from contained stage \bbone\ code into a single expression $p$, 
leaving a residual $l.r$ to interpret the results of that precomputation.
Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
that is, $p$ is a unit value and $l$ is never used in $r$.
At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
the precomputations of the constituent parts, and then unbundling them with a pattern,
with the residual precisely resembling the original expression.

The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subexpression to isolate the precomputation.
It may seem like this $\prev$ rule is throwing information away, but we well show later that the first component 
of the evaluated form of $c$ in the $\prev$ rule must always be unit.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the residual.%
\footnote{The residual of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}

@@

In the splitting rules for stage \bbone\ {\tt if} and {\tt case} expressions, we
have enough information at stage \bbone\ to know what branch to take, so there's
no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
taken branch only, and inject the precomputation into a sum type. In the
residual, we case on that injection, and resume the stage \bbtwo\ portion of the
correct branch.

@@

The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
handle this by splitting the function into two functions: one in the immediate
result which handles all the stage \bbone\ content of the original, and one in
the residual which handles all of the stage \bbtwo\ content. Note that stage
\bbone\ lambdas are partial values, and so have only a trivial precomputation.

@@

We can now develop an notion of what it means for splitting to be correct.  
Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
Thus, we start with two mutually dependent definitions of equivalence.  
Both relate evaluation-residuals on the left with splitting residuals on the right,
but $\equiv$ equates closed expressions, whereas $\sim$ equates values.

\begin{definition}
For evaluation-residual $q$ and splitting-residual $r$, define $q \equiv r : A$ to mean that 
$q \tworedsym v_q$ iff $\reduce {r} {v_r}$ where $v_q \sim v_r : A$, 
\end{definition}

\begin{definition}
For evaluation-residual value $v_1$ and splitting-residual value $v_2$, define $v_1 \sim v_2 : A$ by the following cases:
\begin{itemize}
\item $i \sim i : \rmint$
\item $(v_1,u_1) \sim (v_2,u_2) : A \times B$ where $v_1 \sim v_2 : A$ and $u_1 \sim u_2 : B$
\item $\inl~v_1 \sim \inl~v_2 : A + B$ where $v_1 \sim v_2 : A$
\item $\inr~v_1 \sim \inr~v_2 : A + B$ where $v_1 \sim v_2 : B$
\item $\lam {x_1} A {e_1} \sim \lambda x_2.e_2 : A \to B$ where \\ $\forall (v_1 \sim v_2 : A). [v_1/x_1]e_1 \equiv [v_2/x_2]e_2 : B$
\end{itemize}
\end{definition}

>>>>

Splitting arbitrary terms in \lang\ again results in two terms containing only
the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
called, respectively, the {\em precomputation} and {\em resumer}.
%
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a resumer
\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
the precomputation must agree with the original term. This process of hoisting
all the stage \bbone\ subterms out of a term is called \emph{lifting}.

@@

The precomputation for this term is a \emph{combined result} which is
the pair of these components: \verb|(((),3*4),1+2)|. The resumer is
likewise the pair of the stage \bbtwo\ results of lifting and masking:
\verb|fn L => (L<5,())|.
%\footnote{In this example, the combined result is a 2-tuple, but more generally
%it only must be a term which reduces to a 2-tuple.}

@@

The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
$c$ and resumer $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
precomputation $p$ and a resumer $l.r$. In both cases, the term $e$ is open on
variables in the context $\Gamma$. We write the rules using pattern notation
when appropriate.
%For concision, we represent resumers as a term open on a single variable,
%rather than as functions.

@@

The $\splittwosym$ rules are conceptually simple and almost all very similar.  
As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
from contained stage \bbone\ code into a single expression $p$, 
leaving a resumer $l.r$ to interpret the results of that precomputation.
Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
that is, $p$ is a unit value and $l$ is never used in $r$.
At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
the precomputations of the constituent parts, and then unbundling them with a pattern,
with the resumer precisely resembling the original expression.

The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subexpression to isolate the precomputation.
It may seem like this $\prev$ rule is throwing information away, but we well show later that the first component 
of the evaluated form of $c$ in the $\prev$ rule must always be unit.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the resumer.%
\footnote{The resumer of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}

@@

In the splitting rules for stage \bbone\ {\tt if} and {\tt case} expressions, we
have enough information at stage \bbone\ to know what branch to take, so there's
no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
taken branch only, and inject the precomputation into a sum type. In the
resumer, we case on that injection, and resume the stage \bbtwo\ portion of the
correct branch.

@@

The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
handle this by splitting the function into two functions: one in the immediate
result which handles all the stage \bbone\ content of the original, and one in
the resumer which handles all of the stage \bbtwo\ content. Note that stage
\bbone\ lambdas are partial values, and so have only a trivial precomputation.

@@

We can now develop an notion of what it means for splitting to be correct.  
Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
Thus, we start with two mutually dependent definitions of equivalence.  
Both relate residuals on the left with resumers on the right,
but $\equiv$ equates closed expressions, whereas $\sim$ equates values.

\begin{definition}
For residual $q$ and resumer $r$, define $q \equiv r : A$ to mean that 
$q \tworedsym v_q$ iff $\reduce {r} {v_r}$ where $v_q \sim v_r : A$, 
\end{definition}

\begin{definition}
For residual value $v_1$ and resumer value $v_2$, define $v_1 \sim v_2 : A$ by the following cases:
\begin{itemize}
\item $i \sim i : \rmint$
\item $(v_1,u_1) \sim (v_2,u_2) : A \times B$ where $v_1 \sim v_2 : A$ and $u_1 \sim u_2 : B$
\item $\inl~v_1 \sim \inl~v_2 : A + B$ where $v_1 \sim v_2 : A$
\item $\inr~v_1 \sim \inr~v_2 : A + B$ where $v_1 \sim v_2 : B$
\item $\lam {x_1} A {e_1} \sim \lambda x_2.e_2 : A \to B$ where \\ $\forall (v_1 \sim v_2 : A). [v_1/x_1]e_1 \equiv [v_2/x_2]e_2 : B$
\end{itemize}
\end{definition}
